---
title: "Fitting simple occupancy models in TMB etc."
author: "B. Bolker"
date: '`r Sys.time()`'
output:
  html_document: default
---

# To do

- pull in .cpp/stan/jags from files, display/highlight
- what the hell is Empirical Bayes doing?
- get Stan code working here, try it in "optimize" mode
- no real point trying JAGS ... NIMBLE??
- write tidying methods

# Overview

This is cut down from Olivier Gimenez's dynamic-occupancy equivalent, for my own educational purposes.

Simulate some data (366 sites, 40 samples/site), model using TMB, ADMB, JAGS and Unmarked and finally perform a quick benchmarking. See Olivier's [previous post](https://oliviergimenez.github.io/post/occupancy_in_admb/) (on dynamic occupancy models: here we're cutting it down to simple models).

```{r pkgs, message=FALSE, warning=FALSE}
library(tidyverse)
library(unmarked)
library(TMB)
library(rbenchmark)
library(broom.mixed)
library(Hmisc) ## llist
library(rstan)
```

```{r simulate, message=FALSE, warning=FALSE}
R <- 366 # number of sites
J <- 40  # samples per site
beta_p <- c(-1,1) # fixed effect parameters for detection
beta_phi <- c(-1,1) # fixed effect parameters for occupancy

set.seed(101)
dd <- expand.grid(site=factor(1:R), sample=factor(1:J))
x_occ_0 <- rnorm(R)
occ_prob <- plogis(model.matrix(~x_occ_0) %*% beta_phi)
occ_val <- rbinom(R, size=1, prob=occ_prob)
dd <- (dd
  %>% mutate(x_p = rnorm(R*J),  ## detection covariate
         x_phi = x_occ_0[site], ## occupancy covariate
         occ = occ_val[site],   ## map occupancy to samples
         det_prob = plogis(model.matrix(~x_p) %*% beta_p)*occ,
         obs = rbinom(R*J, size=1, prob=det_prob)
         )
)
```

Back-calculate capture histories and site info for `unmarked::occu`:

```{r unmarked_extract}
## basic site-level info
siteCovs_0 <- (dd
  %>% select(site,x_phi)
  %>% unique()
)
## unmarked::occu() wants it "naked"; no tibbles
siteCovs <- (siteCovs_0
  %>% select(-site)
  %>% as.data.frame()
)
obsCovs <- (dd
  %>% arrange(site) ## need site-major order!
  %>% select(x_p)
  %>% as.data.frame()
)
y <- (dd
  %>% select(site,sample,obs)
  %>% spread(sample,obs)
  %>% select(-site)
  %>% as.matrix()
)
```

```{r unmarked_run}
uf <- unmarkedFrameOccu(y, siteCovs=siteCovs, obsCovs=obsCovs)
(occu_fit <- occu( ~ x_p ~ x_phi, data=uf))
```

For most of the other approaches it will be convenient to construct the model matrices ourselves in R, before them passing them to the relevant toolbox.

```{r make_model_matrices}
dd <- ( dd
  %>% arrange(site)      ## make sure we're in site-major order!
  %>% mutate(n=seq(n())) ## add obs index
)
site_tot <- dd %>% group_by(site) %>% summarise(x=sum(obs)) %>% pull(x)
## indices of beginning *and end* of samples per site
site_ind <- c(0,which(diff(as.numeric(dd$site))>0),
              length(dd$site))
X_phi <- model.matrix(~ x_phi, data=siteCovs_0)
X_p <- model.matrix(~ x_p, data=dd)
## stuff for Stan model
## FIXME: resolve phi/psi differences!
n_site <- nrow(X_phi)
m_psi <- ncol(X_phi)
total_surveys <- nrow(X_p)
m_p <- ncol(X_p)
any_seen <- as.numeric(site_tot>0)
n_survey <- count(dd,site) %>% pull(n) ## or table(dd$site)
start_idx <- dd %>% group_by(site) %>% summarise(n=min(n)) %>% pull(n)
end_idx   <- dd %>% group_by(site) %>% summarise(n=max(n)) %>% pull(n)
```

## TMB

```{r tmb_setup}
compile("SO_TMB.cpp")
dyn.load(dynlib("SO_TMB"))
f <- MakeADFun(
  data = c(as.list(dd[c("obs","site")]),
           list(Xphi = X_phi,
                Xp = X_p,
                site_tot = site_tot,
                site_ind = site_ind)),
  parameters = list(beta_phi = rep(0,ncol(X_phi)),
                    beta_p = rep(0,ncol(X_p))),
  DLL = "SO_TMB",
  silent=TRUE)
TMB_fit <- do.call("optim",f)
class(f) <- "TMB"
broom.mixed::tidy(f)
```

## Stan

```{r stan_build}
sm <- stan_model(file="SO_Stan.stan")
```

```{r stan_opt}
stan_data <- llist(n_site, m_psi, X_psi=X_phi,
                total_surveys, m_p, X_p,
                site=as.integer(dd$site),
                y=dd$obs,
                start_idx, end_idx)
##     data=stan_data)
stan_opt <- optimizing(sm, data=stan_data)
stan_opt$par[1:4]
```

```{r stan_sample, cache=TRUE}
sampling(sm, data=stan_data)
```

```{r bench,cache=TRUE}
(b1 <- benchmark(occu_C=occu( ~ x_p ~ x_phi, data=uf),
                 occu_R=occu( ~ x_p ~ x_phi, data=uf, engine="R"),
                 TMB=do.call("optim",f),
                 stan_opt=optimizing(sm, data=stan_data),
                 replications=5,
                 columns=c("test","elapsed","relative")))
```

FIXME: `microbenchmark` has nicer plots (but tabular output is ugly)

## brain dump

- write out computational details of hierarchical binom model, especially the contortions needed to compute everything in log-space
- formatting options for hierarchical input: capture-history matrix (convenient esp for balanced obs) vs. vector + site indicator (easier when everything is long format)? separate site-level model matrix etc. or combined (inefficient but maybe convenient) site + sample model matrix? Or a list? (The matrix/list minimize indexing to site level ... could provide an index vector ... In any case, need to maintain key distinction that occupancy is modeled at the site level ...)

```{r SI}
sessionInfo()
```
