---
title: "Fitting simple occupancy models in TMB etc."
author: "B. Bolker"
date: '`r Sys.time()`'
output:
  html_document: default
---

# Overview

This is cut down from Olivier Gimenez's dynamic-occupancy equivalent, for my own educational purposes.

Simulate some data (366 sites, 40 samples/site), model using TMB, ADMB, JAGS and Unmarked and finally perform a quick benchmarking. See Olivier's [previous post](https://oliviergimenez.github.io/post/occupancy_in_admb/) (on dynamic occupancy models: here we're cutting it down to simple models).

```{r pkgs, message=FALSE, warning=FALSE}
library(tidyverse)
library(unmarked)
```

```{r message=FALSE, warning=FALSE}
R <- 366 # number of sites
J <- 40  # samples per site
beta_p <- c(-1,1) # fixed effect parameters for detection
beta_psi <- c(-1,1) # fixed effect parameters for occupancy

set.seed(101)
dd <- expand.grid(site=factor(1:R), sample=factor(1:J))
x_occ_0 <- rnorm(R)
occ_prob <- plogis(model.matrix(~x_occ_0) %*% beta_psi)
occ_val <- rbinom(R, size=1, prob=occ_prob)
dd <- dd %>%
  mutate(x_p = rnorm(R*J),      ## detection covariate
         x_psi = x_occ_0[site], ## occupancy covariate
         occ = occ_val[site],   ## map occupancy to samples
         det_prob = plogis(model.matrix(~x_p) %*% beta_p)*occ,
         obs =rbinom(R*J, size=1, prob=det_prob)
         )
```

Back-calculate capture histories and site info for `unmarked::occu`:

```{r unmarked_extract}
## basic site-level info
siteCovs_0 <- (dd
  %>% select(site,x_psi)
  %>% unique()
)
## unmarked::occu() wants it "naked"; no tibbles
siteCovs <- (siteCovs_0
  %>% select(-site)
  %>% as.data.frame()
)
obsCovs <- (dd
  %>% arrange(site) ## need site-major order!
  %>% select(x_p)
  %>% as.data.frame()
)
y <- (dd
  %>% select(site,sample,obs)
  %>% spread(sample,obs)
  %>% select(-site)
  %>% as.matrix()
)
```

```{r unmarked_run}
uf <- unmarkedFrameOccu(y, siteCovs=siteCovs, obsCovs=obsCovs)
occu( ~ x_p ~ x_psi, data=uf)
```

For most of the other approaches it will be convenient to construct the model matrices ourselves in R, before them passing them to the relevant toolbox.

```{r make_model_matrices}
X_psi <- model.matrix(~ x_psi, data=siteCovs_0)
X_p <- model.matrix(~ x_p, data=dd)
```

## brain dump

- formatting options: capture-history matrix (convenient esp for balanced obs) vs. vector + site indicator (easier when everything is long format)? separate site-level model matrix etc. or combined (inefficient but maybe convenient) site + sample model matrix? Or a list? (The matrix/list minimize indexing to site level ... could provide an index vector ... In any case, need to maintain key distinction that occupancy is modeled at the site level ...)

```{r}
# list of data
df = list(K=K,N=N,JJ=JJ,nh=nh,e=garb,data=yy,eff=eff,primary=primary,secondarybis=secondary_bis)
params <- list(logit_psi = 0, logit_det = 0, logit_gam = 0, logit_eps = 0) ## starting parameters

#----- ADMB

library(R2admb)
model <- 
paste("
DATA_SECTION
 init_int K // Number of seasons
 init_int N // Number of seasons x surveys
 init_int JJ // Number of time intervals between primary occasions
 init_int nh // Number of sites
 init_ivector e(1,nh) // Date of first capture
 init_imatrix data(1,nh,1,N) // Data matrix
 init_ivector eff(1,nh) // Number of individuals per capture history
 init_ivector primary(1,K) // seasons
 init_ivector secondarybis(1,JJ) // time intervals between primary occasions
PARAMETER_SECTION
 init_bounded_number logit_psi(-20.0,20.0,1) // init occupancy
 init_bounded_number logit_det(-20.0,20.0,1) // detection
 init_bounded_number logit_gam(-20.0,20.0,1) // colonization
 init_bounded_number logit_eps(-20.0,20.0,1) // extinction
 objective_function_value g
// number psi
// number det
// number gam
// number eps
 sdreport_number psi
 sdreport_number det
 sdreport_number gam
 sdreport_number eps
 PROCEDURE_SECTION
 psi = mfexp(logit_psi);
 psi = psi/(1+psi); 
 det = mfexp(logit_det);
 det = det/(1+det); 
 gam = mfexp(logit_gam);
 gam = gam/(1+gam); 
 eps = mfexp(logit_eps);
 eps = eps/(1+eps); 
 dvar_vector prop(1,2);
 prop(1) = 1-psi; prop(2) = psi;
 dvar_matrix B(1,2,1,2);
 B(1,1) = 1;
 B(1,2) = 1-det;
 B(2,1) = 0.0;
 B(2,2) = det;
 dvar3_array PHI(1,N,1,2,1,2);
 for(int i=1;i<=K;i++){
 	PHI(primary(i),1,1) = 1-gam;
 	PHI(primary(i),1,2) = gam;
 	PHI(primary(i),2,1) = eps;
 	PHI(primary(i),2,2) = 1-eps;
 }
 for(int j=1;j<=JJ;j++){
 	PHI(secondarybis(j),1,1) = 1;
 	PHI(secondarybis(j),1,2) = 0;
 	PHI(secondarybis(j),2,1) = 0;
 	PHI(secondarybis(j),2,2) = 1;
 }

 for(int i=1;i<=nh;i++){
 	int oe = e(i) + 1; // initial obs
 	ivector evennt = data(i)+1; //
 	dvar_vector ALPHA = elem_prod(prop,B(oe));
 	for(int j=2;j<=N;j++){
 		ALPHA = elem_prod(ALPHA*PHI(j-1),B(evennt(j)));
 		g -= log(sum(ALPHA))*eff(i);
 	}
 }
")
writeLines(model,"model.tpl")
setup_admb("/Applications/ADMBTerminal.app/admb")

#---- Unmarked

library(unmarked)
simUMF <- unmarkedMultFrame(y = yy,numPrimary=K)

#---- JAGS

library(jagsUI)
model <- 
paste("
    model{
    #priors
    p ~ dunif(0,1)
    psi1 ~ dunif(0,1)
    epsilon ~ dunif(0,1)
    gamma~dunif(0,1)
   
    for(i in 1:n.sites){
        # process
        z[i,1] ~ dbern(psi1)
        for(t in 2:n.seasons){ 
            mu[i,t]<-((1-epsilon)*z[i,t-1])+(gamma*(1-z[i,t-1]))
            z[i,t]~dbern(mu[i,t])
        }
        # obs
        for(t in 1:n.seasons){ 
            for(j in 1:n.occas){
                p.eff[i,j,t] <- z[i,t]*p
                y[i,j,t]~dbern(p.eff[i,j,t])
            }
        }
    }
}
")
writeLines(model,"dynocc.txt")

jags.data <- list(y=y,n.seasons=dim(y)[3],n.occas=dim(y)[2],n.sites=dim(y)[1])
z.init <- apply(jags.data$y,c(1,3),max)
initial <- function()list(p=runif(1,0,1),psi1=runif(1,0,1),z=z.init,epsilon=runif(1,0,1),gamma=runif(1,0,1))
params.to.monitor <- c("psi1","gamma","epsilon","p")
inits <- list(initial(),initial())

#---- TMB

library(TMB)

model <- 
paste("
#include <TMB.hpp>

/* implement the vector - matrix product */
template<class Type>
vector<Type> multvecmat(array<Type>  A, matrix<Type>  B) {
  int nrowb = B.rows();
  int ncolb = B.cols(); 
  vector<Type> C(ncolb);
  for (int i = 0; i < ncolb; i++)
  {
	    C(i) = Type(0);
      for (int k = 0; k < nrowb; k++){
        C(i) += A(k)*B(k,i);
    }
  }
  return C;
}

template<class Type>
Type objective_function<Type>::operator() () {
  
  // b = parameters
  PARAMETER_VECTOR(b);
  
  // data
  DATA_IMATRIX(ch); // ch = site histories
  DATA_IVECTOR(e); // e = init state 
  DATA_IVECTOR(primary); // seasons
  DATA_IVECTOR(secondarybis); // time intervals between primary occasions 
  
  int nh = ch.rows(); // nb of sites
  int N = ch.cols(); // nb of seasons x surveys
  int JJ = secondarybis.size(); // nb of time intervals between primary occasions
  int K = primary.size(); // nb of seasons
  
  int npar = b.size();

  vector<Type> par(npar);
  for (int i = 0; i < npar; i++) {
    par(i) = Type(1.0) / (Type(1.0) + exp(-b(i)));
  }
  Type psi = par(0); // init occupancy
  Type det = par(1); // detection
  Type gam = par(2); // col
  Type eps = par(3); // ext
  // careful, indexing starts at 0!
  
  // init states - occ
  vector<Type> PROP(2);
  PROP(0) = Type(1.0)-psi;
  PROP(1) = psi;

  // obs
  matrix<Type> B(2,2);
  B(0,0) = Type(1.0);
  B(0,1) = Type(1.0) - det;
  B(1,0) = Type(0.0);
  B(1,1) = det;

  // transition
  array<Type> PHI(2,2,N);
  for (int i = 0; i < K; i++) {
    PHI(0,0,primary(i)) = Type(1.0) - gam;
    PHI(0,1,primary(i)) = gam;
    PHI(1,0,primary(i)) = eps;
    PHI(1,1,primary(i)) = Type(1.0) - eps;
  }
  
  for (int j = 0; j < JJ; j++) {
    PHI(0,0,secondarybis(j)) = Type(1.0);
    PHI(0,1,secondarybis(j)) = Type(0.0);
    PHI(1,0,secondarybis(j)) = Type(0.0);
    PHI(1,1,secondarybis(j)) = Type(1.0);
  }

  REPORT(PHI);
  
  // likelihood
  Type ll(0);
  Type nll(0);
  array<Type> ALPHA(2);
  for (int i = 0; i < nh; i++) {
    vector<int> evennt = ch.row(i);
    ALPHA = PROP * vector<Type>(B.row(e(i))); // element-wise vector product
    REPORT(ALPHA);
    for (int j = 1; j < N; j++) {
      matrix<Type> TEMP = PHI.col(j-1);
      matrix<Type> PHIj(2,2);
      PHIj(0,0) = TEMP(0,0);
      PHIj(0,1) = TEMP(1,0);
      PHIj(1,0) = TEMP(2,0);
      PHIj(1,1) = TEMP(3,0);
      ALPHA = multvecmat(ALPHA,PHIj)* vector<Type>(B.row(evennt(j))); // vector matrix product, then element-wise vector product
    }
    ll += log(sum(ALPHA));
  }
  nll = -ll;
  return nll;
}
")
writeLines(model,"occ_tmb.cpp")

# compile
compile("occ_tmb.cpp")

# load
dyn.load(dynlib("occ_tmb"))

# inits
binit = rep(0,4)

# match model/data
f <- MakeADFun(
  data = list(ch = yy, e = garb, primary=primary-1,secondarybis=secondary_bis-1), 
  parameters = list(b = binit),
  DLL = "occ_tmb")

## fit the model (optimization)
#opt <- do.call("optim", f)
#opt
#sdreport(f) # get SEs

#----- benchmarking!

library(microbenchmark)
res = microbenchmark(
do_admb('model', data=df, params = params),
colext(psiformula= ~1, gammaformula = ~ 1, epsilonformula = ~ 1,pformula = ~ 1, data = simUMF, method="BFGS",se=FALSE),
jagsUI(data=jags.data, inits, parameters.to.save=params.to.monitor, model.file="dynocc.txt",n.thin = 1,n.chains = 2, n.burnin = 500, n.iter =1000,parallel=TRUE),
do.call("optim", f),times=3)
res2 = summary(res)
res2
```

The results are amazing! TMB is `r res2$median[1]/res2$median[4]` faster than ADMB, `r res2$median[2]/res2$median[4]` than Unmarked and `r res2$median[3]/res2$median[4]` faster than Jags. 

# Conclusions

I'm new to TMB, but I'm gonna definitely dig into it. Congrats to the developers!
